A state space model reveals the rich and varied dynamics of hippocampal replay

### 2020-07-21 Article.tex
Update on Overleaf.

Hippocampal replay events are a prototypical example of a generative representation. As animals move through space, neurons in the hippocampus preferentially fire in at specific locations in an environment, and thus sets of cells can fire in sequence as the animal moves through a series of locations. When the animal is asleep or immobile, hippocampal cells can be reactivated during a "sharp-wave ripple" (SWR) event. A subset of SWRs contain sequential firing similar to that seen during a previous or potential future experience, and previous work has reported that these sequential firing events proceed at an average speed of ~10 meters / second, about 20x faster than the animal's usual movement speed \cite{DavidsonHippocampalReplayExtended2009}.

Identifying these events typically involves multiple steps and assumptions about the nature of the event. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on an threshold for event size chosen by the experimenter \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. The spikes within these events decode the probability of position in overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach to detecting sequential replay involves fitting a line to the resulting set of probability distributions, which instantiates the assumption that the representations progress at a constant speed \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009}. A statistical test is then used to determining whether the fit is better than the fit to shuffled versions of the data.

While this approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bin acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic, because it has the potential to reject real events that do not move at constant speeds. Approaches that focus on the order of cell activity within each even \cite{LeeMemorySequentialExperience2002, GuptaHippocampalReplayNot2010} allow for a relaxation of that linear assumption, but replace by either ignoring all but the first spike from each cell or some arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, there is now evidence that a subset of these events have stationary representations of a single location, which both approaches exclude. In addition, these methods identify only a subset of events as containing content (typically somewhere between 10 and 45\% of the chosen events), and thereby provide little insight into the remaining majority of the events.

Recognizing the problems with the linear fit, several studies have moved away from the constant velocity assumption, using the weighted mean or maximum of each posterior time bin and connecting each time bin with a line. For example, using this approach, Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015} found that replays can alternate between representing a single location and sequential spatial trajectories. On the other hand, Stella et al. \cite{StellaHippocampalReactivationRandom2019} found that replays are more spatially continuous, following Brownian diffusion dynamics. Both methods still used large time bins and neither took into account the uncertainty of the decoded estimates, however, making it hard to identify the source of the different conclusions.

An ideal approach to identifying and quantifying the dynamics of generative representations would circumvent these problems. It would use all of the available spiking data to yield the most accurate decoded positions. It would be applicable to either thresholded events or to all of the data to permit an unbiased assessment of the prevalence and nature of generative activity. It would use very small temporal bins (1 or 2 ms) to allow for very rapid representational movement and provide information about the certainty of the decoded estimates. It would be able to capture different types of movement dynamics, ranging from stationary to rapidly evolving to disorganized, and provide a statistical assessment of confidence for the dynamic. Finally, where assumptions are made, it would provide well defined parameters whose values can be explored systematically to understand their influence on the results.

We therefore developed a state space model of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \cite{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} and characterizes represented trajectories as a mixture of three underlying patterns of dynamics: stationary trajectories, continuous trajectories that are many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike wave form features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to data from 10 animals, focusing on SWR events to permit a direct comparison to previous work.

@@

Finally, we specify how likely each dynamic is to persist in time versus changing to another dynamic (Figure \ref{1}D). In order to be conservative, we assume that each dynamic is likely to dominate for ~100 ms, which is approximately the duration of the SWR. There is, however, a small probability of switching to one of the other dynamics (note that below we show that our results are relatively insensitive to this parameter). Accordingly, we set the probability of staying in a dynamic to 0.98 for each 2 ms time step. Because the probability of changing dynamics follows a geometric distribution, this corresponds to an expected duration of 100 ms for staying in a particular dynamic. Importantly, the data drive the dynamic, so even if the probability of a particular dynamic is high, data providing clear evidence for a change in dynamic can, within the time frame of ~10 ms, drive a change to a different dynamic.

@@

As mentioned above, the model is robust to the choice of probability of staying in a state. To demonstrate this we decoded the spiking sequence in Figure \ref{1}A with different probabilities of staying in the same state versus switching to another state (Figure \ref{fig:Figure1-Figure supplement 1}). We found that for a large range of plausible probabilities of staying in one of the dynamics (0.9 - 0.9999), the model still correctly identified the dynamic with high probability.

@@

As expected, and providing further support for the validity of the method, we observed many replays that were classified as continuous throughout the entirety of SWR (Figure \ref{fig:Figure2-Figure supplement 2}A-C). However, we also observed many events with spatially coherent content that did not have this structure. For example, there were trajectories that started in one direction and reversed back to the original position (Figure \ref{3}A, Figure \ref{fig:Figure3-Figure supplement 1}1), trajectories that remained fixed in one position (Figure \ref{3}B), trajectories that jumped between arms and between dynamics (Figure \ref{3}C-D, Figure \ref{fig:Figure3-Figure supplement 1}A, E, F), and trajectories that were spatially incoherent throughout the SWR (Figure \ref{fig:Figure3-Figure supplement 1}D).

@@

\subsection*{Position of the Animal and Linearization}
The animal's 2D position was estimated from digital video (30 Hz) of two infrared diodes placed on the head stage preamplifiers using a semi-automated analysis. In order to decrease the time it takes to run the model, the 2D position of the animal was converted into a 1D position. This is done by first defining a 2D graph representation of the track (herein referred to as the track graph), where edges correspond to segments of the W-track and nodes represent intersection points between those segments. Then, based on the algorithm in \cite{NewsonHiddenMarkovmap2009}, we use a Hidden Markov Model (HMM) to classify the most likely track segment the animal is on. Using the HMM takes into account the time dependent nature of the data and helps prevents sudden jumps from one track segment to another, which is particularly important near intersections. The observation model of the HMM is Gaussian and it models the likelihood of being on a track segment as the Gaussian distance to that segment with a 5 cm standard deviation. The state transition model is an empirically estimated state transition that changes with each time point that tries to ensure that the Euclidean distance between successive position estimates is similar to the shortest path distance along the graph between successive position estimates. A slight bias of 0.1 is given to the diagonal to encourage staying on the same track segment. The most likely track segment the animal is on is computed using the Viterbi algorithm. After finding the track segment that corresponds to each 2D position, the 2D position is projected onto the track segment. This allows us to define a distance from the center well in terms of shortest path length on the track, where 0 cm represents the center well position. The linear distance can then be converted into a linear position by assigning each track segment a position in 1D space. 15 cm gaps were placed between the center arm, left arm, and right arms in 1D space to prevent any smoothing done in the model from bleeding over inappropriately. The code used for linearization can be found at \url{https://github.com/Eden-Kramer-Lab/loren_frank_data_processing}.

@@

\subsection*{The Model}
Let $x_{k}$ be a continuous latent variable that corresponds to the position represented by the population of cells at time $t_k$ and $I_{k}$ be a discrete latent variable that is an indicator for the dynamics we wish to characterize: stationary, continuous, and fragmented. The goal of the model is to simultaneously estimate the posterior probability of position and dynamics $p(x_k, I_k \mid O_{1:T})$, where $O_{1:T}$ corresponds to the observed spiking data from time 1 to time $T$. The observed data can be either spike trains $\Delta N_{1:T}^{(1:C)}$ from $C$ putative cells when decoding with sorted spikes or multiunit spikes $\Delta N_{1:T}^{(1:E)}$ and their associated wave form features $\Vec{m}$ from each tetrode $E$ when decoding with clusterless spikes.

@@

\subsection*{Encoding - Clusterless}
In order to relate each tetrode's unsorted spiking activity and waveform features to position, we used kernel density estimation (KDE) to estimate the following distributions: $\pi(x)$, $p_{i}(x_k)$, and $p_{i}(x_k, \vec{m}_k)$. We used KDEs of the form: 
$$
kde(y) = \frac{1}{N h_1...h_D} \sum^{N}_{i=1} \prod^{D}_{d=1} K_{d}\left(\frac{y_d - y_{i,d}}{h_d}\right)
$$
where $y$ is the data with $D$ dimensions, $K$ is a 1 dimensional Gaussian kernel with bandwidth $h_d$, $y_{i, d}$ is the data observed during movement, $N$ is the number of observations during the movement period. For $\pi(x)$, $y_{i, d}$ is all positions observed during movement. For $p_{i}(x_k)$, $y_{i, d}$ is all positions at the time of multiunit spikes during movement. For $p_{i}(x_k, \vec{m}_k)$, $y_{i, d}$ is all positions at the time of multiunit spikes and their associated waveform features during movement. We choose the bandwidth $h_d$ to be 6.0 cm for any position dimension and 24.0 $\mu V$ for any spike amplitude dimension, because these parameters minimized decoding error in \cite{KloostermanBayesiandecodingusing2014}. The mean firing rate $\mu_k$ was also estimated during movement.

@@

\subsection*{Highest Posterior Density}
The 95\% highest posterior density (HPD) is a measurement of the spread of the posterior probability and is defined as the region of the posterior that contains the top 95\% values of the posterior probability. In this manuscript, we use the HPD to evaluate the uncertainty of the posterior probability of position. We calculate the 95\% HPD by determining the maximum threshold value $h$ that fulfills the follow equality:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} p(x_{k} \mid O_{1:T})dx = 0.95
$$
The 95\% HPD is the set of position bins with posterior values greater than the threshold $\{x : p(x_{k} \mid O_{1:T}) > h\}$. The cumulative spatial extent of the HPD is calculated by taking the integral of the members of this set:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} \mathbb{1}dx
$$
which we approximate with a Reimann sum.

>>>>

Hippocampal replay events are a prototypical example of a generative representation. As animals move through space, neurons in the hippocampus preferentially fire at specific locations in an environment, and thus sets of cells can fire in sequence as the animal moves through a series of locations. When the animal is asleep or immobile, hippocampal cells can be reactivated during a "sharp-wave ripple" (SWR) event. A subset of SWRs contain sequential firing similar to that seen during a previous experience or likely to occur during potential future experiences, and previous work has reported that these sequential firing events proceed at an average speed of ~10 meters / second, about 20x faster than the animal's usual movement speed \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

Identifying these events typically involves multiple steps and assumptions about their nature. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. The spikes within these events decode the probability of position in overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach to detecting sequential replay involves fitting a line to the resulting set of probability distributions, which instantiates the assumption that the representations progress at a constant speed \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the fit is better than the fit to shuffled versions of the data.

While this approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bin acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic, because it has the potential to reject real events that do not move at constant speeds. Approaches that focus on the order of cell activity within each even \cite{LeeMemorySequentialExperience2002, GuptaHippocampalReplayNot2010} allow for a relaxation of that linear assumption, but replace it with a loss of statistical power due to either ignoring all but the first spike from each cell or using some arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, there is now evidence that a subset of these events have stationary representations of a single location, which both approaches often exclude. In addition, these methods identify only a subset of events as containing content (typically somewhere between 10 and 45\% of the chosen events), and thereby provide little insight into the remaining majority of the events.

Recognizing the problems with the linear fit, several studies have moved away from the constant velocity assumption, using an estimated represented location at each time bin and connecting each with a line. For example, using this approach, Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015} found that replays can alternate between representing a single location and sequential spatial trajectories. On the other hand, Stella et al. \cite{StellaHippocampalReactivationRandom2019} found that replays are more spatially continuous, following Brownian diffusion dynamics. Both methods still used large time bins and neither took into account the uncertainty of the decoded estimates, however, making it hard to identify the source of the different conclusions.

An ideal approach to identifying and quantifying the dynamics of generative representations would circumvent these problems. It would use all of the available spiking data to yield the most accurate decoded positions. It would be applicable to either thresholded events or to all of the data to permit an unbiased assessment of the prevalence and nature of generative activity. It would use very small temporal bins (1 or 2 ms) to allow for very rapid representational movement and provide information about the certainty of the decoded estimates. It would be able to capture different types of movement dynamics, ranging from stationary to rapidly evolving to disorganized, and provide a statistical assessment of confidence for each category of dynamics. Finally, where assumptions are made, it would provide well defined parameters whose values can be explored systematically to understand their influence on the results.

We therefore developed a state space model of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \cite{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} and characterizes represented trajectories as a mixture of three underlying patterns of movement dynamics: stationary trajectories, continuous trajectories that are many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike waveform features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to data from 10 rats, focusing on SWR events to permit a direct comparison to previous work.

@@

Finally, we specify how likely each dynamic is to persist in time versus changing to another dynamic (Figure \ref{1}D). In order to be conservative, we assume that each dynamic is likely to dominate for \texttildelow100 ms, which is approximately the duration of the SWR. There is, however, a small probability of switching to one of the other dynamics (note that below we show that our results are relatively insensitive to this parameter). Accordingly, we set the probability of staying in a dynamic to 0.98 for each 2 ms time step. Because the probability of changing dynamics follows a geometric distribution, this corresponds to an expected duration of 100 ms for staying in a particular dynamic. Importantly, the data drives the estimated dynamic, so even if the probability of a particular dynamic is high, data providing clear evidence for a change in dynamic can, within the time frame of \texttildelow10 ms, drive a change to a different estimated dynamic.

@@

As mentioned above, the model is robust to the choice of probability of staying in a state. To demonstrate this we decoded the spiking sequence in Figure \ref{1}A with different probabilities of staying in the same state versus switching to another state (Figure \ref{fig:Figure1-Figure supplement 1}). We found that for a large range of plausible probabilities of staying in one of the dynamics over a single 2 ms time step (0.9 - 0.9999), the model still correctly identified the dynamic with high probability.

@@

As expected, and providing further support for the validity of the method, we observed many replays that were classified as continuous throughout the entirety of SWR (Figure \ref{fig:Figure2-Figure supplement 2}A-C). However, we also observed many events with spatially coherent content that did not have this structure. For example, there were trajectories that started in one direction and reversed back to the original position (Figure \ref{3}A, Figure \ref{fig:Figure3-Figure supplement 1}), trajectories that remained fixed in one position (Figure \ref{3}B), trajectories that jumped between arms and between dynamics (Figure \ref{3}C-D, Figure \ref{fig:Figure3-Figure supplement 1}A, E, F), and trajectories that were spatially incoherent throughout the SWR (Figure \ref{fig:Figure3-Figure supplement 1}D).

@@

\subsection*{Position of the Animal and Linearization}
The animal's 2D position was estimated from digital video (30 Hz) of two infrared diodes placed on the head stage preamplifiers using a semi-automated analysis. In order to decrease the time it takes to run the model, the 2D position of the animal was converted into a 1D position. This is done by first defining a 2D graph representation of the track (herein referred to as the track graph), where edges correspond to segments of the W-track and nodes represent intersection points between those segments. Then, based on the algorithm in \cite{NewsonHiddenMarkovmap2009}, we use a Hidden Markov Model (HMM) to classify the most likely track segment the animal is on. Using the HMM takes into account the time dependent nature of the data and helps prevents sudden jumps from one track segment to another, which is particularly important near intersections. The observation model of the HMM is Gaussian and it models the likelihood of being on a track segment as the Gaussian distance to that segment with a 5 cm standard deviation. The state transition model is an empirically estimated state transition that changes with each time point that tries to ensure that the Euclidean distance between successive position estimates is similar to the shortest path distance along the graph between successive position estimates. A slight bias of 0.1 is given to the diagonal to encourage staying on the same track segment. The most likely track segment the animal is on is computed using the Viterbi algorithm. After finding the track segment that corresponds to each 2D position, the 2D position is projected onto the nearest point of the track segment. This allows us to define a distance from the center well in terms of shortest path length on the track, where 0 cm represents the center well position. The linear distance can then be converted into a linear position by assigning each track segment a position in 1D space. 15 cm gaps were placed between the center arm, left arm, and right arms in 1D space to prevent any smoothing done in the model from bleeding over inappropriately. The code used for linearization can be found at \url{https://github.com/Eden-Kramer-Lab/loren_frank_data_processing}.

@@

\subsection*{The Model}
Let $x_{k}$ be a continuous latent variable that corresponds to the position represented by the population of cells at time $t_k$ and $I_{k}$ be a discrete latent variable that is an indicator for the dynamics we wish to movement characterize: stationary, continuous, and fragmented. The goal of the model is to simultaneously estimate the posterior probability of position and dynamics $p(x_k, I_k \mid O_{1:T})$, where $O_{1:T}$ corresponds to the observed spiking data from time 1 to time $T$. The observed data can be either spike trains $\Delta N_{1:T}^{(1:C)}$ from $C$ putative cells when decoding with sorted spikes or multiunit spikes $\Delta N_{1:T}^{(1:E)}$ and their associated wave form features $\Vec{m}$ from each tetrode $E$ when decoding with clusterless spikes.

@@

\subsection*{Encoding - Clusterless}
In order to relate each tetrode's unsorted spiking activity and waveform features to position, we used kernel density estimation (KDE) to estimate the following distributions: $\pi(x)$, $p_{i}(x_k)$, and $p_{i}(x_k, \vec{m}_k)$. We used KDEs of the form: 
$$
kde(y) = \frac{1}{N h_1...h_D} \sum^{N}_{i=1} \prod^{D}_{d=1} K_{d}\left(\frac{y_d - y_{i,d}}{h_d}\right)
$$
where $y$ is the data with $D$ dimensions, $K$ is a 1 dimensional Gaussian kernel with bandwidth $h_d$, $y_{i, d}$ is the data observed during movement, $N$ is the number of observations during the movement period. For $\pi(x)$, $y_{i, d}$ is all positions observed during movement. For $p_{i}(x_k)$, $y_{i, d}$ is all positions at the time of multiunit spikes during movement. For $p_{i}(x_k, \vec{m}_k)$, $y_{i, d}$ is all positions at the time of multiunit spikes and their associated waveform features during movement. We choose the bandwidth $h_d$ to be 6.0 cm for any position dimension and 24.0 $\mu V$ for any spike amplitude dimension, because these parameters were found to minimize decoding error in Kloosterman et al. \cite{KloostermanBayesiandecodingusing2014}. The mean firing rate $\mu_k$ was also estimated during movement.

@@

\subsection*{Highest Posterior Density}
The 95\% highest posterior density (HPD) is a measurement of the spread of the posterior probability and is defined as the region of the posterior that contains the top 95\% values of the posterior probability \cite{CasellaStatisticalinference2001}. In this manuscript, we use the HPD to evaluate the uncertainty of the posterior probability of position. We calculate the 95\% HPD by determining the maximum threshold value $h$ that fulfills the follow equality:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} p(x_{k} \mid O_{1:T})dx = 0.95
$$
The 95\% HPD is the set of position bins with posterior values greater than the threshold $\{x : p(x_{k} \mid O_{1:T}) > h\}$. The cumulative spatial extent of the HPD is calculated by taking the integral of the members of this set:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} \mathbb{1}dx
$$
which we approximate with a Reimann sum.
